---
title: "Biostat 203B Homework 1 Solution"
author: Liqiao (Vicky) Li
subtitle: Due Jan 22 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=F}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email.

**Solution**: Done.

2. Create a **private** repository `biostat-203b-2021-winter` and add `Hua-Zhou`, `Chris-German` and `ElvisCuiHan` as your collaborators with write permission.

**Solution**: Done.

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `master` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `master` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and data sets to reproduce results) in `master` branch.

**Solution**: Done.

4. After each homework due date, teaching assistant and instructor will check out your master branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

**Solution**: Done.

5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

**Solution**: Done.

## Q2. Linux Shell Commands

1. This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic-iv.mit.edu/docs/access/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)

**Solution**: Here is my [course completion report](https://www.citiprogram.org/verify/?ka7f91ad8-1de0-400a-b550-c7f2b8169102-40353745) and [course completion certificate](https://www.citiprogram.org/verify/?wfcbbdf07-618a-4efa-9d7d-81e9d2bda5ab-40353745).

2. The `/usr/203b-data/mimic-iv/` folder on teaching server contains data sets from MIMIC-IV. Refer to <https://mimic-iv.mit.edu/docs/datasets/> for details of data files.  
    ```{bash}
    ls -l /usr/203b-data/mimic-iv
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/usr/203b-data/mimic-iv` directly in following exercises. 

    Use Bash commands to answer following questions.

3. Display the contents in the folders `core`, `hosp`, `icu`. What are the functionalities of the bash commands `zcat`, `zless`, `zmore`, and `zgrep`? 

**Solution**: 
    ```{bash}
   ls -l /usr/203b-data/mimic-iv/core
    ```

    ```{bash}
   ls -l /usr/203b-data/mimic-iv/hosp
    ```

    ```{bash}
   ls -l /usr/203b-data/mimic-iv/icu
    ```
    
Here are the functionalities of the bash commands:
`zcat`: view the contents of a compressed file (i.e., gz file)
`zless`: view a compressed file one page at a time, scroll upwards and downwards through the input
`zmore`: view a gz file one page at a time (only downwards)
`zgrep`: search out expressions from a given compressed gz file

4. What's the output of following bash script?
    ```{bash, eval=F}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        ls -l $datafile
      done
    ```
Display the number of lines in each data file using a similar loop.

**Solution**: 
The output of the bash script above is the list of detailed contents of the directory `/usr/203b-data/mimic-iv/core/`.

Below are the number of lines in each data file in folder `core`:
    ```{bash, eval=F}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        wc -l $datafile
      done
    ```
Below are the number of lines in each data file in folder `hosp`:
    ```{bash, eval=F}
    for datafile in /usr/203b-data/mimic-iv/hosp/*.gz
      do
        wc -l $datafile
      done
    ```

Below are the numbers of lines in each data file in folder `icu`:
    ```{bash, eval=F}
    for datafile in /usr/203b-data/mimic-iv/icu/*.gz
      do
        wc -l $datafile
      done
    ```

5. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, `language`, `marital_status`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq`, `wc`, and so on.)

**Solution**: 

The first 10 lines of `admissions.csv.gz`:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |head -10
    ```

The number of rows in `admissions.csv.gz`is 524521:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | awk 'END { print NR}'
    ```

The number of unique patients in this data file is 257366:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |awk -F, '{ print $1 }'|
    tail -n +2 | sort | uniq | wc -l

    ```

**The possible values taken by each variable are listed below:**

admission_type:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |awk -F, '{ print $6 }'|
    tail -n +2 | sort | uniq
    ```
    
admission_location:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |awk -F, '{ print $7 }'|
    tail -n +2 | sort | uniq
    ```

insurance:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |awk -F, '{ print $9 }'|
    tail -n +2 | sort | uniq
    ```

language:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |awk -F, '{ print $10 }'|
    tail -n +2 | sort | uniq
    ```

marital_status:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |awk -F, '{ print $11 }'|
    tail -n +2 | sort | uniq
    ```

ethnicity:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |awk -F, '{ print $12 }'|
    tail -n +2 | sort | uniq
    ```


**The count for each unque value of these variables are listed below:**

The number of each unique value of `admission_type`is 9:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |awk -F, '{ print $6 }'|
    tail -n +2 | sort | uniq | wc -l
    ```

The number of each unique value of `admission_location`is 12:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |awk -F, '{ print $7 }'|
    tail -n +2 | sort | uniq | wc -l
    ```

The number of each unique value of `insurance`is 3:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |awk -F, '{ print $9 }'|
    tail -n +2 | sort | uniq | wc -l
    ```

The number of each unique value of `language`is 2:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |awk -F, '{ print $10 }'|
    tail -n +2 | sort | uniq | wc -l
    ```
    
The number of each unique value of `marital_status`is 5:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |awk -F, '{ print $11 }'|
    tail -n +2 | sort | uniq | wc -l
    ```

The number of each unique value of `ethnicity`is 8:
    ```{bash}
    zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz |awk -F, '{ print $12 }'|
    tail -n +2 | sort | uniq | wc -l
    ```
## Q3. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=FALSE}
    curl http://www.gutenberg.org/cache/epub/42671/pg42671.txt > pride_and_prejudice.txt
    ```
Do **not** put this text file `pride_and_prejudice.txt` in Git. Using a `for` loop, how would you tabulate the number of times each of the four characters is mentioned?

**Solution**:
    ```{bash}
    for name in 'Elizabeth' 'Jane' 'Lydia' 'Darcy'
      do
      echo $name
      grep -o $name pride_and_prejudice.txt | wc -l
      
    done
    ```

2. What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```

**Solution**: The first command directs output to display "hello, world" in the file test1.txt and overwrite previously existing scripts already in test1.txt; The second command appends output to test2.txt, so "hello, world" will be added to test2.txt without deleting the previously existing scripts, if any.

3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=FALSE}
    ./middle.sh pride_and_prejudice.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

**Solution**: I use commands `touch middle.sh` and `vi middle.sh` to create the 'middle.sh' file. Then I use `chmod`make the file executable by the owner and run the commands below   

```{bash, eval=FALSE}
chmod u+x middle.sh 
./middle.sh pride_and_prejudice.txt 20 5
```

The output displays lines 15-20 of the Pride and Prejudice.txt file. A same output can be obtained by running the commands below: 
```{bash, eval=FALSE}
head -n 20 pride_and_prejudice.txt | tail -n 5
```

Thus, `"$1"` in the shell scripts represents the file name, `"$2"` represents the number of lines from the top of the text file, and `"3"` represents the number of lines from the bottom to be displayed. The first lineof the shell scripts tells the system to use the share `/bin/sh` to run these commands.

### Q4. More fun with Linux

Try these commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.

**Solution**: `cal` displays the calendar of current month:
```{bash}
cal
```
`cal 2021` displays the calendar of 2021:
```{bash}
cal 2021
```
`cal 9 1752` displays the calendar of September in 1752:
```{bash}
cal 9 1752
```
`date` displays the current date and time of the system:
```{bash}
date
```
`hostname` displays the name of the current teaching server:
```{bash}
hostname
```
`arch` displays the current instruction set architecture and system version:
```{bash}
arch
```
`uname -a` displays Linux system information:
```{bash}
uname -a
```
`uptime` displays how long the system has been running and load:
```{bash}
uptime
```
`who am i` displays who you are logged in as:
```{bash}
whoami
```
`who` displays who is logged into the system:
```{bash}
who
```
`w` displays who is online:
```{bash}
w
```
`id` displays the user and group ids of your current user：
```{bash}
id
```
`last | head` displays the first 10 lines of the last users who have logged onto the system:
```{bash}
last | head
```
`echo {con,pre}{sent,fer}{s,ed}` displays all possible combinations of characters from three groups: 
```{bash}
echo {con,pre}{sent,fer}{s,ed}
```
`time sleep 5` displays the time taken to pause the script for 5 seconds:
```{bash}
time sleep 5
```
`history | tail` displays the last 10 previously executed command:
```{bash}
set -o history
echo "hello"
history | tail
```
