---
title: "Biostat 203B Homework 2 Solution"
subtitle: Due Feb 5 @ 11:59PM
author name: "Liqiao (Vicky) Li"
output: 
  html_document:
    toc: true
    toc_depth: 4 
---

Display machine information for reproducibility:
```{r}
sessionInfo()
```

```{r setup}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(tidyverse)
library(data.table)
library(lubridate)
```

```{r}
os <- sessionInfo()$running
if (str_detect(os, "Linux")) {
  mimic_path <- "/usr/203b-data/mimic-iv"
} else if (str_detect(os, "macOS")) {
  mimic_path <- "/Users/huazhou/Documents/Box Sync/MIMIC/mimic-iv-0.4"
}
```

Use tidyverse (ggpot2, dplyr) to explore the [MIMIC-IV](https://mimic-iv.mit.edu) data introduced in [homework 1](https://ucla-biostat203b-2021winter.github.io/hw/hw1/hw1.html).

```{r}
system(str_c("tree -s -L 2 ", shQuote(mimic_path)), intern = TRUE)
```

## Q1. PhysioNet credential

At this moment, you should already get credentialed on the PhysioNet. Please include a screenshot of your `Data Use Agreement for the MIMIC-IV (v0.4)`.

**Solution**: I have not got credentialed on the PhysioNet yet.

## Q2. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package. Is there any speed difference?

In this homework, we stick to the tidyverse. 

**Solution**:
```{r}
admission <- read.csv(str_c(mimic_path, "/core/admissions.csv.gz"))

```

```{r}
admission <- read_csv(str_c(mimic_path, "/core/admissions.csv.gz"))
```

```{r}
admission <- fread(str_c(mimic_path,"/core/admissions.csv.gz"))

```

Among the three programs, `read.csv` is the slowest,`read_csv` is much faster than `read.csv`, and `fread` is even faster than `read_csv`.

## Q3. ICU stays

`icustays.csv.gz` (<https://mimic-iv.mit.edu/docs/datasets/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. Summarize following variables using appropriate numerics or graphs:   

- how many unique `stay_id`?  
- how many unique `subject_id`?  
- length of ICU stay  
- first ICU unit  
- last ICU unit  

**Solution**:

```{r}
icustay <- read_csv(str_c(mimic_path, "/icu/icustays.csv.gz")) %>%
                      print(width = Inf)
```


Check number of unique `stay_id`. It seems each row is a unique ICU stay. The number of unique `stay_id` is 69619: 
```{r}
icustay %>%
  distinct(stay_id) %>%
  nrow()
```

The number of unique `subject_id`is 50048, showing that each patient can have multiple stays:
```{r}
icustay %>%
  distinct(subject_id) %>%
  nrow()
```

Length of ICU stay:
```{r}
icustay %>%
  ggplot(aes(x = los)) +
  geom_boxplot(mapping = aes(x = factor(""), y = los)) +
  scale_y_log10()+
  labs(x = "Length of ICU Stays", y = "Days")

summarise(icustay, los_quantile = quantile(los, c(0.25, 0.5, 0.75)))
```
As shown in the box plot, the length of ICU stays has a wide range. The median (IQR) length of ICU stay is 2.06 (1.13 - 4.20) days.


First ICU unit:
```{r}
icustay %>%
  ggplot() +
  geom_bar(mapping = aes(x = first_careunit, fill = first_careunit)) +
  theme(axis.text.x = element_blank()) +
  labs(x = "First Care Unit", y = "Number of ICU Stays")

icustay %>%
  group_by(first_careunit) %>%
  summarise(
    count = n()
  )
```
There are 11 first care units. Most patients initially stayed in MICU, MICU/SICU, CVICU, and SICU which have more than 10,000 stays. In contrast, only few stays occurred in Medicine and PACU.


Last ICU stay:
```{r}
icustay %>%
  ggplot() +
  geom_bar(mapping = aes(x = last_careunit, fill = last_careunit)) +
  theme(axis.text.x = element_blank()) +
  labs(x = "Last Care Unit", y = "Number of ICU stays")

icustay %>%
  group_by(last_careunit) %>%
  summarise(
    count = n()
  )
```

There are 9 last care units in total. Similar to first care unit, MICU, MICU/SICU, CVICU, and SICU are the most popular units.

## Q4. `admission` data

Information of the patients admitted into hospital is available in `ADMISSION.csv.gz`. See <https://mimic-iv.mit.edu/docs/datasets/core/admissions/> for details of each field in this file. Summarize following variables using appropriate graphs. Explain any patterns you observe.   

- admission year  
- admission month  
- admission month day  
- admission week day  
- admission hour (anything unusual?)  
- number of deaths in each year  
- admission type  
- number of admissions per patient  
- admission location  
- discharge location  
- insurance  
- language  
- martial status  
- ethnicity  
- death 

Note it is possible that one patient (uniquely identified by the `SUBJECT_ID`) is admitted into hospital multiple times. When summarizing some demographic information, it makes sense to summarize based on unique patients. 

## Q5. `patient` data

Explore `patients.csv.gz` (<https://mimic-iv.mit.edu/docs/datasets/core/patients/>) and summarize following variables using appropriate numerics and graphs:  

- `gender`  
- `anchor_age` (explain pattern you see)

## Q6. Lab results

`labevents.csv.gz` (<https://mimic-iv.mit.edu/docs/datasets/hosp/labevents/>) contains all laboratory measurements for patients. 

We are interested in the lab measurements of creatinine, potassium, sodium, chloride, bicarbonate, hematocrit, white blood cell count, glucose, magnesium, calcium, phosphorus, and lactate. Find the `itemid`s of these lab measurements from `d_labitems.csv.gz` and retrieve a subset of `labevents.csv.gz` only containing these items.

## Q7. Vitals from chartered events

We are interested in the vitals for ICU patients: heart rate, mean and systolic blood pressure (invasive and noninvasive measurements combined), body temperature, SpO2, and respiratory rate.

`chartevents.csv.gz` (<https://mimic-iv.mit.edu/docs/datasets/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `ITEMID` variable indicates a single measurement type in the database. The `VALUE` variable is the value measured for `ITEMID`. 

`d_items.csv.gz` (<https://mimic-iv.mit.edu/docs/datasets/icu/d_items/>) is the dictionary for the `ITEMID` in `CHARTEVENTS.csv`. Find potential values of `ITEMID` that correspond to systolic blood pressure, i.e., `LABEL` contains the string `systolic`. 

## Q8. Putting things together

Let us create a tibble for all ICU stays, where rows are  

- first ICU stay of each unique patient  
- adults (age at admission > 18)  

and columns contains at least following variables  

- all variables in `icustays.csv.gz`  
- all variables in `admission.csv.gz`  
- first lab measurements during ICU stay  
- first vitals measurement during ICU stay  
- an indicator variable whether the patient died within 30 days of hospital admission  

